---
title: "Problem Set 1"
author: "Nate GOnzales-Hess"
date: "2025-04-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
library(tidyverse)  # Load any packages you need here
```

## Problem 1: Globe Tossing Posterior Distribution

*Suppose the globe tossing experiment turned out to be 3 water and 11 land. Construct the posterior distribution.*

```{r problem1}
# Function to compute posterior for a given sample and range of possible proportions.
compute_posterior = function(sample, possible){
  
  W = sum(sample == "W")
  L = sum(sample == "L")
  
  prior = rep(1, length(possible)) # 2. Calculate likelihood and prior
  likelihood = sapply(possible, function(x) dbinom(x = W, size = W+L, prob = x))
  post = (prior * likelihood) / sum(prior * likelihood)  # 3. Calculate posterior probability
  
  return(post)
}

# Sample with 3 water and 13 land:
sample = c("W","W","W","L","L","L","L","L","L","L","L","L","L","L","L","L")
# 1. Create a grid of possible proportion values
possible = seq(0,1,length.out=50)
# Compute posteriors
posterior = compute_posterior(sample, possible)

# 4. Create a plot of the posterior distribution
barplot(posterior, names.arg = round(possible,2),
        main = "Posterior probabilities for possible proportions of water to land",
        xlab = "Proportion Water",
        ylab = "Density")
```
Your interpretation here:



## Problem 2: Posterior Predictive Distribution

*Using the posterior distribution from Problem 1, compute the posterior predictive distribution for the next 5 tosses of the globe.*

```{r problem2}
# Number of steps in our simulation
n_tosses = 5
sim_iters = 1000
# Empty vector to store predicted "W" counts (0-5 possible "W" observations)
predicted_counts = rep(0, n_tosses+1)

# 1. Sample from your posterior distribution
post_samples <- sample(possible, size=sim_iters, prob=posterior, replace=TRUE )

# 2. Use these samples to simulate new tosses
sim_globe = function( p=0.7 , N=5 ){
  sample(
    x = c("W", "L"),  # possible values
    size = N,         # how many draws
    prob = c(p, 1-p), # probability of each possibility
    replace = TRUE    # the same value can be drawn multiple times
  )
}

# Run simulation
for (i in 1:sim_iters){
  p = post_samples[i]
  tosses = sim_globe(p, n_tosses)
  water_count = sum(tosses == "W")
  predicted_counts[water_count+1] = predicted_counts[water_count+1] + 1
}

# 3. Create a visualization of the predictions
barplot(predicted_counts, names.arg = seq(0,5),
        main = "Posterior Predictive Distribution (5 new tosses)",
        xlab = "Number of \"W\"",
        ylab = "Count")

```

Your interpretation here:


## Session Information

```{r session-info}
# This will help with debugging and reproducibility
sessionInfo()
```