---
title: "Problem Set 4"
author: "Nate Gonzales-Hess"
date: "2025-04-28"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rethinking)
library(brms)
library(here)
library(tinytex)
library(ggdist)
```

## Run Models from lecture 3-1:
```{r}
# Load data and filter
d <- rethinking::sim_happiness(seed = 1990, N_years = 1000)
rethinking::precis(d)

d2 <- d[d$age >= 18, ]
d2$A <- rethinking::standardize(d2$age)
d2$mid <- as.factor(d2$married + 1)

# Model in which happiness is influenced by both marriage and age
m6a <- brm(
  data=d2, 
  family=gaussian,
  bf( happiness ~ 0 + a + b*A, 
      a ~ 0 + mid,
      b ~ 0 + mid,
      nl = TRUE),
  prior = c( prior(normal(0, .50), nlpar=a),
             prior(normal(0, .25), nlpar=b),
             prior(exponential(1), class=sigma)),
  iter=2000, warmup=1000, seed=9, chains=1,
  file = here("files/models/31.6a")
)

# Model in which happiness is only influenced by age
m7 <- brm(
  data=d2, 
  family=gaussian,
  happiness ~ A,
  prior = c( prior(normal(0, .50), class=Intercept),
             prior(normal(0, .25), class=b),
             prior(exponential(1), class=sigma)),
  iter=2000, warmup=1000, seed=9, chains=1,
  file = here("files/models/31.7")
)
```

## Apply PSIS and WAIC to models:

```{r}
# WAIC
waic(m6a)
waic(m7)
```
```{r}
waic_comparison <- loo_compare(waic(m6a), waic(m7))
print(waic_comparison)
```
```{r}
# PSIS
loo(m6a)
loo(m7)
```
```{r}
# Compare models
psis_comparison <- loo_compare(loo(m6a), loo(m7))
print(psis_comparison)
```
### Interpretation:
WAIC and Loo give nearly identical results, and both indicate that model 6a (the more complex model) performs better in terms of predictive accuracy than model 7. Despite superior predictive performance, model 6a actually gets the causal relationship wrong, as when we stratify by marriage (Lecture 3-1, slide 39), we see that age is the more causal variable.

## We’re Number One, Alas:
```{r}
data("Laffer")

# Simple linear model (straight line)
model_linear <- brm(
  formula = tax_revenue ~ tax_rate,
  data = Laffer,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = "Intercept"),
    prior(normal(0, 10), class = "b"),
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  cores = 4,
  file = here("files/models/ps4_linear")
)

# Quadratic model (curve)
model_quadratic <- brm(
  formula = tax_revenue ~ tax_rate + I(tax_rate^2),
  data = Laffer,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = "Intercept"),
    prior(normal(0, 10), class = "b"),
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  cores = 4,
  file = here("files/models/ps4_quadratic")
)
```

```{r}
# Create smooth prediction data
pred_data <- data.frame(tax_rate = seq(min(Laffer$tax_rate), max(Laffer$tax_rate), length.out = 100))

# Get the FITTED values directly (rather than draws from the posterior)
# This will give us smooth lines
fitted_linear <- fitted(model_linear, newdata = pred_data)
fitted_quadratic <- fitted(model_quadratic, newdata = pred_data)

# Add fitted values to new_data
pred_data$linear_fit <- fitted_linear[,"Estimate"]
pred_data$quadratic_fit <- fitted_quadratic[,"Estimate"]

# Create plot with smooth lines
ggplot(Laffer, aes(x = tax_rate, y = tax_revenue)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_line(data = pred_data, aes(y = linear_fit, color = "Linear"), size = 1) +
  geom_line(data = pred_data, aes(y = quadratic_fit, color = "Quadratic"), size = 1) +
  scale_color_manual(values = c("Linear" = "blue", "Quadratic" = "red")) +
  labs(title = "Linear vs Quadratic Model for Laffer Data",
       x = "Tax Rate",
       y = "Tax Revenue",
       color = "Model") +
  theme_minimal()
```
```{r warning=FALSE, message=FALSE}
loo_linear = loo(model_linear)
loo_quad = loo(model_quadratic)
```

```{r warning=FALSE, message=FALSE}
curve_comparison <- loo_compare(loo(model_linear), loo(model_quadratic))
print(curve_comparison)
```
### Interpretation:
The quadratic model seems to have slightly better prediction accuracy than the linear model, however because the se_diff is so close to the elpd_diff, I don't think there is any practical difference in prediction quality between the two models.

## Outlier identification using Pareto k (We already know it's Norway, but.):
```{r}
# Extract Pareto k values
pareto_k_linear <- loo_linear$diagnostics$pareto_k
pareto_k_quad <- loo_quad$diagnostics$pareto_k

# Identify observations with high k values
high_k_linear <- which(pareto_k_linear > 0.7)
high_k_quad <- which(pareto_k_linear > 0.7)
```

```{r message=FALSE, warning=FALSE}
# Make models without Norway datapoint
no_norway <- Laffer[-high_k_linear,]

# Simple linear model (straight line)
model_linear_no_norway <- brm(
  formula = tax_revenue ~ tax_rate,
  data = no_norway,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = "Intercept"),
    prior(normal(0, 10), class = "b"),
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  cores = 4,
  file = here("files/models/ps4_linear_no_norway")
)

# Quadratic model (curve)
model_quadratic_no_norway <- brm(
  formula = tax_revenue ~ tax_rate + I(tax_rate^2),
  data = no_norway,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = "Intercept"),
    prior(normal(0, 10), class = "b"),
    prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  cores = 4,
  file = here("files/models/ps4_quadratic_no_norway")
)
```

```{r}
# Compare Linear
loo(model_linear)
loo(model_linear_no_norway)
```

```{r}
# Compare Quadratic
loo(model_quadratic)
loo(model_quadratic_no_norway)
```

```{r}
post4 = as_draws_df(model_quadratic) %>% mutate(model="With Norway")
post5 = as_draws_df(model_quadratic_no_norway) %>% mutate(model="Without Norway")
full_join(post4, post5) %>% 
  ggplot(aes(x = b_tax_rate, fill = model)) +  # Specify x and fill aesthetics
  stat_halfeye(alpha = 0.8) +
  scale_fill_manual(values = c("grey", "#1c5253")) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Tax Rate Coefficient",
       y = NULL,
       title = "Posterior Distributions With and Without Norway") +
  theme_minimal()
```

### Interpretation:
Removing Norway improves model quality, in particular it reduces standard error. And removing Norway has a regularizing effect on the posterior distribution as a whole.